The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-06-09 manuscript.Rmd
edits to manuscript

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood, and modeling the impact of human activities is limited to a few taxonomic groups. A major hindrance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016]. There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), and difficult to access. These factors have contributed to the lack of synthesis and have made the identification of knowledge gaps difficult. However, with the development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can better evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur in the continental United States, the Gulf of Mexico, and the Pacific Northwest, to highlight gaps in our knowledge. We show that public databases are relatively incomplete, and that most species are known from a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively, representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species. For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig. 1).

```{r compare_database_overlap, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). (\\#fig:database-overlap)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

@@

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text manipulation, it is increasingly easier to validate and spell check the scientific names of the identifications of museum specimens. Additionally, because WoRMS links the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing to compare specimens identified at different time periods.

Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them. While tools have been developed to convert names of locales into GPS coordinates on land, similar approaches may not work for locales at sea. However, geographic information is available in other forms, and can be digitized.

>>>>

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010]. A major hindrance for marine species, is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016]. There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur in the continental United States, the Gulf of Mexico, and the Pacific Northwest, to highlight gaps in our knowledge. We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively, representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species. For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. 1). Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). (\\#fig:database-overlap)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list.', ech=FALSE}
knitr::include_graphics("figures/database_overlap.pdf")
```

@@

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes. Numbers next to the bars indicate the number of species.'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text manipulation, it is increasingly easier to validate and spell check the scientific names of the identifications of museum specimens. Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability. Additionally, because species names registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods, a tedious process that would have required expert knowledge in the past.

Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea. However, geographic information is available in other forms, and can be digitized.
